---
title: "Intermediate R"
output: 
 html_document:
   toc: true
   toc_depth: 2
   number_sections: true
   toc_float: true
---

```{css, include=FALSE}
.badCode {
background-color: red;
}
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Conditionals and Control Flow
## Relational Operators

  - Equality `==`
  - Inequality `!=`
  - `<` and `>`
  - `<=` and `>=`
```{r}
# Comparison of logicals
TRUE == FALSE

# Comparison of numerics
-6 * 14 != 17 - 101

# Comparison of character strings
"useR" == "user"

# Compare a logical with a numeric
TRUE == 1

# Comparison of numerics
-6 * 5 + 2 >= -10 + 1

# Comparison of character strings
"raining" <= "raining dogs"

# Comparison of logicals
TRUE > FALSE

# The linkedin and facebook vectors have already been created for you
linkedin <- c(16, 9, 13, 5, 2, 17, 14)
facebook <- c(17, 7, 5, 16, 8, 13, 14)

# Popular days
linkedin > 15

# Quiet days
linkedin <= 5

# LinkedIn more popular than Facebook
linkedin > facebook

views <- matrix(c(linkedin, facebook), nrow = 2, byrow = TRUE)

# When does views equal 13?
views == 13

# When is views less than or equal to 14?
views <= 14
```
## Logical Operators

  - AND operator "&"
  - OR operator "|"
  - NOT operator "!"
  - "&" vs "&&", "|" vs "||" (the double ampersand operation only examines the first element of each vector.)

```{r}
# The linkedin and last variable are already defined for you
linkedin <- c(16, 9, 13, 5, 2, 17, 14)
last <- tail(linkedin, 1)

# Is last under 5 or above 10?
last <5 | last >10

# Is last between 15 (exclusive) and 20 (inclusive)?
last > 15 & last <=20

# linkedin exceeds 10 but facebook below 10
linkedin > 10 & facebook < 10

# When were one or both visited at least 12 times?
linkedin >= 12 | facebook >=12

# When is views between 11 (exclusive) and 14 (inclusive)?
views > 11 & views <= 14
```
## Conditional Statements
> if (condition) {
  expr
}

>if (condition) {
  expr1
} else {
  expr2
}

```{r}
# Variables related to your last day of recordings
li <- 15
fb <- 9

# Code the control-flow construct
if (li >= 15 & fb >= 15) {
  sms <- 2 * (li + fb)
} else if (li < 10 & fb < 10) {
  sms <- 0.5 * (li + fb)
} else {
  sms <- li + fb
}

# Print the resulting sms to the console
sms
```


# Loops

# Functions

# The apply family

# Utilities

---
This code was last compiled at `r Sys.time()`