---
title: "Introduction to R"
output: 
 html_document:
   toc: true
   toc_depth: 2
   number_sections: true
   toc_float: true
---

```{css, include=FALSE}
.badCode {
background-color: red;
}
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Intro to basics

R makes use of the # sign to add comments, so that you and others can understand what the R code is about. Comments are not run as R code, so they will not influence your result. 
```{r}
# Calculate 3 + 4
3 + 4

# Calculate 6 + 12
6 + 12
```
In its most basic form, R can be used as a simple calculator. Consider the following arithmetic operators:

  - Addition: `+`
  - Subtraction: `-`
  - Multiplication: `*`
  - Division: `/`
  - Exponentiation: `^`
  - Modulo: `%%`
  
The last two might need some explaining:

  - The `^` operator raises the number to its left to the power of the number to its right: for example 3^2 is 9.
  - The modulo returns the remainder of the division of the number to the left by the number on its right, for example 5 modulo 3 or 5 %% 3 is 2.
```{r}
# An addition
5 + 5 

# A subtraction
5 - 5 

# A multiplication
3 * 5

 # A division
(5 + 5) / 2 

# Exponentiation
2^5

# Modulo
28%%6
```
A basic concept in (statistical) programming is called a **variable**.

A variable allows you to store a value (e.g. 4) or an object (e.g. a function description) in R. You can then later use this variable's name to easily access the value or the object that is stored within this variable.
```{r}
# Assign the value 42 to x
x <- 42

# Print out the value of the variable x
x

# Assign a value to the variables my_apples and my_oranges
my_apples <- 5
my_oranges <- 6

# Add these two variables together
my_apples + my_oranges

# Create the variable my_fruit
my_fruit <- my_apples+my_oranges
```
R works with numerous data types. Some of the most basic types to get started are:

  - Decimal values like 4.5 are called numerics.
  - Whole numbers like 4 are called integers. Integers are also numerics.
  - Boolean values (TRUE or FALSE) are called logical.
  - Text (or string) values are called characters.
  
# Vectors

Vectors are one-dimension arrays that can hold numeric data, character data, or logical data. In other words, a vector is a simple tool to store data. For example, you can store your daily gains and losses in the casinos.

In R, you create a vector with the combine function c(). You place the vector elements separated by a comma between the parentheses.
```{r}
numeric_vector <- c(1, 10, 49)
character_vector <- c("a", "b", "c")

# Complete the code for boolean_vector
boolean_vector <- c(TRUE,FALSE,TRUE)

# Poker winnings from Monday to Friday
poker_vector <- c(140, -50, 20, -120, 240)

# Roulette winnings from Monday to Friday
roulette_vector <-  c(-24, -50, 100, -350, 10)

# Assign days as names of poker_vector
names(poker_vector) <- c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday")

# Assign days as names of roulette_vector
names(roulette_vector) <- c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday")

# The variable days_vector
days_vector <- c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday")
 
# Assign the names of the day to roulette_vector and poker_vector
names(poker_vector) <- days_vector
names(roulette_vector) <- days_vector

A_vector <- c(1, 2, 3)
B_vector <- c(4, 5, 6)

# Take the sum of A_vector and B_vector
total_vector <- A_vector + B_vector
  
# Print out total_vector
total_vector

# Assign to total_daily how much you won/lost on each day
total_daily <- poker_vector + roulette_vector

# Total winnings with poker
total_poker <- sum(poker_vector)

# Total winnings with roulette
total_roulette <-  sum(roulette_vector)

# Total winnings overall
total_week <- sum(poker_vector,roulette_vector)

# Print out total_week
total_week
  
# Check if you realized higher total gains in poker than in roulette
total_poker > total_roulette
```
To select elements of a vector (and later matrices, data frames, …), you can use square brackets. Between the square brackets, you indicate what elements to select.
```{r}
# Define a new variable based on a selection
poker_wednesday <- poker_vector[3]

poker_midweek <- poker_vector[c(2,3,4)]

roulette_selection_vector <- roulette_vector[2:5]

# Select poker results for Monday, Tuesday and Wednesday
poker_start <- poker_vector[c("Monday", "Tuesday", "Wednesday")]
  
# Calculate the average of the elements in poker_start
mean(poker_start)
```
The (logical) comparison operators known to R are:

  - `<` for less than
  - `>` for greater than
  - `<=` for less than or equal to
  - `>=` for greater than or equal to
  - `==` for equal to each other
  - `!=` not equal to each other

```{r}
# Which days did you make money on poker?
selection_vector <- poker_vector >0
  
# Print out selection_vector
selection_vector

# Select from poker_vector these days
poker_winning_days <- poker_vector[selection_vector]

# Which days did you make money on roulette?
selection_vector <- roulette_vector > 0

# Select from roulette_vector these days
roulette_winning_days <- roulette_vector[selection_vector]
```

# Matrices

In R, a matrix is a collection of elements of the same data type (numeric, character, or logical) arranged into a fixed number of rows and columns. Since you are only working with rows and columns, a matrix is called two-dimensional.

You can construct a matrix in R with the matrix() function. Consider the following example:

> matrix(1:9, byrow = TRUE, nrow = 3)

In the matrix() function:

  - The first argument is the collection of elements that R will arrange into the rows and columns of the matrix. Here, we use 1:9 which is a shortcut for c(1, 2, 3, 4, 5, 6, 7, 8, 9).
  - The argument byrow indicates that the matrix is filled by the rows. If we want the matrix to be filled by the columns, we just place byrow = FALSE.
  - The third argument nrow indicates that the matrix should have three rows.
  
```{r}
# Construct a matrix with 3 rows that contain the numbers 1 up to 9
matrix(1:9, byrow = TRUE, nrow = 3)

# Box office Star Wars (in millions!)
new_hope <- c(460.998, 314.4)
empire_strikes <- c(290.475, 247.900)
return_jedi <- c(309.306, 165.8)

# Create box_office
box_office <- c(new_hope, empire_strikes, return_jedi)

# Construct star_wars_matrix
star_wars_matrix <- matrix(box_office, byrow = TRUE, nrow = 3)

# Vectors region and titles, used for naming
region <- c("US", "non-US")
titles <- c("A New Hope", "The Empire Strikes Back", "Return of the Jedi")

# Name the columns with region
colnames(star_wars_matrix) <- region

# Name the rows with titles
rownames(star_wars_matrix) <- titles

# Print out star_wars_matrix
star_wars_matrix

# Calculate worldwide box office figures
worldwide_vector <- rowSums(star_wars_matrix)
worldwide_vector

# Bind the new variable worldwide_vector as a column to star_wars_matrix
all_wars_matrix <- cbind(star_wars_matrix, worldwide_vector)
all_wars_matrix

star_wars_matrix2 <- matrix(c(474.5,  552.5, 310.7,  338.7, 380.3,  468.5),byrow = TRUE,nrow = 3)
colnames(star_wars_matrix2) <- region
rownames(star_wars_matrix2) <- c("The Phantom Menace", "Attack of the Clones", "Revenge of the Sith")

# Combine both Star Wars trilogies in one matrix
all_wars_matrix <- rbind(star_wars_matrix, star_wars_matrix2)
all_wars_matrix

# Total revenue for US and non-US
total_revenue_vector <- colSums(all_wars_matrix)
  
# Print out total_revenue_vector
total_revenue_vector


```
Similar to vectors, you can use the square brackets [ ] to select one or multiple elements from a matrix. Whereas vectors have one dimension, matrices have two dimensions. You should therefore use a comma to separate the rows you want to select from the columns. For example:

  - `my_matrix[1,2]` selects the element at the first row and second column.
  - `my_matrix[1:3,2:4]` results in a matrix with the data on the rows 1, 2, 3 and columns 2, 3, 4.

If you want to select all elements of a row or a column, no number is needed before or after the comma, respectively:

  - `my_matrix[,1]` selects all elements of the first column.
  - `my_matrix[1,]` selects all elements of the first row.
  
```{r}
# Select the non-US revenue for all movies
non_us_all <- all_wars_matrix[,2]
  
# Average non-US revenue
mean(non_us_all)
  
# Select the non-US revenue for first two movies
non_us_some <- all_wars_matrix[1:2,2]
  
# Average non-US revenue for first two movies
mean(non_us_some)
```
Similar to what you have learned with vectors, the standard operators like `+`, `-`, `/`, `*`, etc. work in an element-wise way on matrices in R.

For example, `2 * my_matrix` multiplies each element of my_matrix by two.
```{r}
# Estimate the visitors
visitors <- all_wars_matrix / 5
  
# Print the estimate to the console
visitors

ticket_prices_matrix <- matrix(c(5,5,6,6,7,7,4,4,4.5,4.5,4.9,4.9),byrow = TRUE,nrow = 6)
rownames(ticket_prices_matrix) <- row.names(all_wars_matrix)
colnames(ticket_prices_matrix) <- colnames(all_wars_matrix)
ticket_prices_matrix

# Estimated number of visitors
visitors <- all_wars_matrix / ticket_prices_matrix

# US visitors
us_visitors <- visitors[,1]

# Average number of US visitors
mean(us_visitors)
```
  
# Factors

The term factor refers to a statistical data type used to store categorical variables. The difference between a categorical variable and a continuous variable is that a categorical variable can belong to a **limited number of categories**. A continuous variable, on the other hand, can correspond to an infinite number of values.

It is important that R knows whether it is dealing with a continuous or a categorical variable, as the statistical models you will develop in the future treat both types differently. (You will see later why this is the case.)

A good example of a categorical variable is sex. In many circumstances you can limit the sex categories to "Male" or "Female". (Sometimes you may need different categories. For example, you may need to consider chromosomal variation, hermaphroditic animals, or different cultural norms, but you will always have a finite number of categories.)

o create factors in R, you make use of the function factor(). First thing that you have to do is create a vector that contains all the observations that belong to a limited number of categories. For example, sex_vector contains the sex of 5 different individuals:

> sex_vector <- c("Male","Female","Female","Male","Male")

It is clear that there are two categories, or in R-terms 'factor levels', at work here: "Male" and "Female".

The function `**factor()**` will encode the vector as a factor:

> factor_sex_vector <- factor(sex_vector)

```{r}
# Sex vector
sex_vector <- c("Male", "Female", "Female", "Male", "Male")

# Convert sex_vector to a factor
factor_sex_vector <- factor(sex_vector)

# Print out factor_sex_vector
factor_sex_vector
```
There are two types of categorical variables: a nominal categorical variable and an ordinal categorical variable.

A nominal variable is a categorical variable without an implied order. This means that it is impossible to say that 'one is worth more than the other'. For example, think of the categorical variable animals_vector with the categories "Elephant", "Giraffe", "Donkey" and "Horse". Here, it is impossible to say that one stands above or below the other. (Note that some of you might disagree ;-) ).

In contrast, ordinal variables do have a natural ordering. Consider for example the categorical variable temperature_vector with the categories: "Low", "Medium" and "High". Here it is obvious that "Medium" stands above "Low", and "High" stands above "Medium".

```{r}
# Animals
animals_vector <- c("Elephant", "Giraffe", "Donkey", "Horse")
factor_animals_vector <- factor(animals_vector)
factor_animals_vector

# Temperature
temperature_vector <- c("High", "Low", "High","Low", "Medium")
factor_temperature_vector <- factor(temperature_vector, order = TRUE, levels = c("Low", "Medium", "High"))
factor_temperature_vector
```
When you first get a dataset, you will often notice that it contains factors with specific factor levels. However, sometimes you will want to change the names of these levels for clarity or other reasons. R allows you to do this with the function levels():

levels(factor_vector) <- c("name1", "name2",...)
A good illustration is the raw data that is provided to you by a survey. A common question for every questionnaire is the sex of the respondent. Here, for simplicity, just two categories were recorded, "M" and "F". (You usually need more categories for survey data; either way, you use a factor to store the categorical data.)

> survey_vector <- c("M", "F", "F", "M", "M")

Recording the sex with the abbreviations "M" and "F" can be convenient if you are collecting data with pen and paper, but it can introduce confusion when analyzing the data. At that point, you will often want to change the factor levels to "Male" and "Female" instead of "M" and "F" for clarity.

**Watch out**: the order with which you assign the levels is important. If you type levels(factor_survey_vector), you'll see that it outputs [1] "F" "M". If you don't specify the levels of the factor when creating the vector, R will automatically assign them alphabetically. To correctly map "F" to "Female" and "M" to "Male", the levels should be set to c("Female", "Male"), in this order.

```{r}
# Code to build factor_survey_vector
survey_vector <- c("M", "F", "F", "M", "M")
factor_survey_vector <- factor(survey_vector)

# Specify the levels of factor_survey_vector
levels(factor_survey_vector) <- c("Female", "Male")

factor_survey_vector

# Generate summary for survey_vector
summary(survey_vector)

# Generate summary for factor_survey_vector
summary(factor_survey_vector)

# Male
male <- factor_survey_vector[1]

# Female
female <- factor_survey_vector[2]

# Battle of the sexes: Male 'larger' than female?
male > female

# Create speed_vector
speed_vector <- c("medium", "slow", "slow", "medium", "fast")
```
speed_vector should be converted to an ordinal factor since its categories have a natural ordering. By default, the function factor() transforms speed_vector into an unordered factor. To create an ordered factor, you have to add two additional arguments: ordered and levels.

> factor(some_vector,
       ordered = TRUE,
       levels = c("lev1", "lev2" ...))
       
By setting the argument ordered to TRUE in the function factor(), you indicate that the factor is ordered. With the argument levels you give the values of the factor in the correct order.

```{r}
# Convert speed_vector to ordered factor vector
factor_speed_vector <- factor(speed_vector, ordered = TRUE, levels = c("slow", "medium", "fast"))

# Print factor_speed_vector
factor_speed_vector
summary(factor_speed_vector)

# Factor value for second data analyst
da2 <-factor_speed_vector[2]

# Factor value for fifth data analyst
da5 <-factor_speed_vector[5]

# Is data analyst 2 faster than data analyst 5?
da2 > da5
```


# Data frames

You may remember from the chapter about matrices that all the elements that you put in a matrix should be of the same type. Back then, your dataset on Star Wars only contained numeric elements.

When doing a market research survey, however, you often have questions such as:

  - 'Are you married?' or 'yes/no' questions (`logical`)
  - 'How old are you?' (`numeric`)
  - 'What is your opinion on this product?' or other   -   'open-ended' questions (`character`)
  - …
  
The output, namely the respondents' answers to the questions formulated above, is a dataset of different data types. You will often find yourself working with datasets that contain different data types instead of only one.

```{r}
mtcars
```
Working with large datasets is not uncommon in data analysis. When you work with (extremely) large datasets and data frames, your first task as a data analyst is to develop a clear understanding of its structure and main elements. Therefore, it is often useful to show only a small part of the entire dataset.

So how to do this in R? Well, the function head() enables you to show the first observations of a data frame. Similarly, the function tail() prints out the last observations in your dataset.

Both head() and tail() print a top line called the 'header', which contains the names of the different variables in your dataset.
```{r}
# Call head() on mtcars
head(mtcars)

# Call tail() on mtcars
tail(mtcars)
```
Another method that is often used to get a rapid overview of your data is the function str(). The function str() shows you the structure of your dataset. For a data frame it tells you:

  - The total number of observations (e.g. 32 car types)
  - The total number of variables (e.g. 11 car features)
  - A full list of the variables names (e.g. mpg, cyl … )
  - The data type of each variable (e.g. num)
  - The first observations
  
Applying the str() function will often be the first thing that you do when receiving a new dataset or data frame. It is a great way to get more insight in your dataset before diving into the real analysis.
```{r}
# Investigate the structure of mtcars
str(mtcars)
```

```{r}
# Definition of vectors
name <- c("Mercury", "Venus", "Earth", 
          "Mars", "Jupiter", "Saturn", 
          "Uranus", "Neptune")
type <- c("Terrestrial planet", 
          "Terrestrial planet", 
          "Terrestrial planet", 
          "Terrestrial planet", "Gas giant", 
          "Gas giant", "Gas giant", "Gas giant")
diameter <- c(0.382, 0.949, 1, 0.532, 
              11.209, 9.449, 4.007, 3.883)
rotation <- c(58.64, -243.02, 1, 1.03, 
              0.41, 0.43, -0.72, 0.67)
rings <- c(FALSE, FALSE, FALSE, FALSE, TRUE, TRUE, TRUE, TRUE)

# Create a data frame from the vectors
planets_df <- data.frame(name, type, diameter, rotation, rings)

# Check the structure of planets_df
str(planets_df)

# Print out diameter of Mercury (row 1, column 3)
planets_df[1,3]

# Print out data for Mars (entire fourth row)
planets_df[4,]

# Select first 5 values of diameter column
planets_df[1:5,"diameter"]

# Select the rings variable from planets_df
rings_vector <- planets_df[,"rings"]
  
# Print out rings_vector
rings_vector

# Adapt the code to select all columns for planets with rings
planets_df[rings_vector,]
```
Now, let us move up one level and use the function subset(). You should see the subset() function as a short-cut to do exactly the same as what you did in the previous exercises.

> subset(my_df, subset = some_condition)

The first argument of subset() specifies the dataset for which you want a subset. By adding the second argument, you give R the necessary information and conditions to select the correct subset.

The code below will give the exact same result as you got in the previous exercise, but this time, you didn't need the rings_vector!

> subset(planets_df, subset = rings)

```{r}
# Select planets with diameter < 1
subset(planets_df,diameter<1)
```
In data analysis you can sort your data according to a certain variable in the dataset. In R, this is done with the help of the function order().

order() is a function that gives you the ranked position of each element when it is applied on a variable, such as a vector for example:

> a <- c(100, 10, 1000)

> order(a)

> [1] 2 1 3

10, which is the second element in a, is the smallest element, so 2 comes first in the output of order(a). 100, which is the first element in a is the second smallest element, so 1 comes second in the output of order(a).

This means we can use the output of order(a) to reshuffle a:

> a[order(a)]

> [1]   10  100 1000

```{r}
# Use order() to create positions
positions <-  order(planets_df$diameter)

# Use positions to sort planets_df
planets_df[positions,]
```

# Lists

Congratulations! At this point in the course you are already familiar with:

  - Vectors (one dimensional array): can hold numeric, character or logical values. The elements in a vector all have the same data type.
  - Matrices (two dimensional array): can hold numeric, character or logical values. The elements in a matrix all have the same data type.
  - Data frames (two-dimensional objects): can hold numeric, character or logical values. Within a column all elements have the same data type, but different columns can be of different data type.

A list in R is similar to your to-do list at work or school: the different items on that list most likely differ in length, characteristic, and type of activity that has to be done.

A list in R allows you to gather a variety of objects under one name (that is, the name of the list) in an ordered way. These objects can be matrices, vectors, data frames, even other lists, etc. It is not even required that these objects are related to each other in any way.

You could say that a list is some kind super data type: you can store practically any piece of information in it!

```{r}
# Vector with numerics from 1 up to 10
my_vector <- 1:10 

# Matrix with numerics from 1 up to 9
my_matrix <- matrix(1:9, ncol = 3)

# First 10 elements of the built-in data frame mtcars
my_df <- mtcars[1:10,]

# Construct list with these different elements:
my_list <- list(my_vector, my_matrix, my_df)
```
Just like on your to-do list, you want to avoid not knowing or remembering what the components of your list stand for. That is why you should give names to them:

> my_list <- list(name1 = your_comp1, 
                name2 = your_comp2)
                
This creates a list with components that are named name1, name2, and so on. If you want to name your lists after you've created them, you can use the names() function as you did with vectors. The following commands are fully equivalent to the assignment above:

> my_list <- list(your_comp1, your_comp2)

> names(my_list) <- c("name1", "name2")

```{r}
# Adapt list() call to give the components names
my_list <- list(vec = my_vector, mat = my_matrix, df = my_df)

# Print out my_list
my_list

# Finish the code to build shining_list
shining_list <- list(moviename = "The Shining",
                     actors = c("Jack Nicholson",   "Shelley Duvall",   "Danny Lloyd",      "Scatman Crothers",
"Barry Nelson" ),
reviews = data.frame(scores = c(4.5,4,5),sources = c("IMDb1"                     ,"IMDb2" ,"IMDb3"                 ),comments = c("Best Horror Film I Have Ever Seen","A truly brilliant and scary film from Stanley Kubrick","A masterpiece of psychological horror")))
```
Your list will often be built out of numerous elements and components. Therefore, getting a single element, multiple elements, or a component out of it is not always straightforward.

One way to select a component is using the numbered position of that component. For example, to "grab" the first component of shining_list you type

> shining_list[[1]]

A quick way to check this out is typing it in the console. Important to remember: to select elements from vectors, you use single square brackets: [ ]. Don't mix them up!

You can also refer to the names of the components, with [[ ]] or with the $ sign. Both will select the data frame representing the reviews:

> shining_list[["reviews"]]

> shining_list$reviews

Besides selecting components, you often need to select specific elements out of these components. For example, with shining_list[[2]][1] you select from the second component, actors (shining_list[[2]]), the first element ([1]). When you type this in the console, you will see the answer is Jack Nicholson.

```{r}
# Print out the vector representing the actors
shining_list[["actors"]]

# Print the second element of the vector representing the actors
shining_list[["actors"]][2]
```

```{r}
movie_title <- "The Departed"

movie_actors <- c("Leonardo DiCaprio", "Matt Damon",        "Jack Nicholson",   
"Mark Wahlberg",     "Vera Farmiga",      "Martin Sheen" )

# Use the table from the exercise to define the comments and scores vectors
scores <- c(4.6, 5, 4.8, 5, 4.2)
comments <- c("I would watch it again", "Amazing!", "I liked it", "One of the best movies", "Fascinating plot")

# Save the average of the scores vector as avg_review
avg_review <- mean(scores)

# Combine scores and comments into the reviews_df data frame
reviews_df <- data.frame(scores,comments)

# Create and print out a list, called departed_list
departed_list <- list(movie_title,movie_actors,reviews_df,avg_review)
departed_list
```



---
This code was last compiled at `r Sys.time()`